\LARGE Untangling the link between traits, size and growth rate in plants

### 2016-09-06 ms/ms.tex
Refinements of text and figures

We assume growth is fundamentally driven by energy income and subsequent distribution throughout the plant. According to a standard model, the amount of biomass available for growth, ${\rm d}B / {\rm d}t$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000} (Fig. \ref{fig:conceptual}a, eqn \ref{eq:dbdt}). Photosynthesis is proportional to the average photosynthetic rate per unit leaf area $\bar{p}$ and total leaf area, $A_{\rm l} = M_{\rm l} / \phi$, where $\phi$ is leaf mass per leaf area. The constant $y$ in eqn \ref{eq:dbdt} converts mol CO$_2$ into units of biomass. We assume that  that $\bar{p}$ increases with light but with diminishing returns, and respiration and turnover rates of different tissues are constants that might differ by trait (see below). While many vegetation models use a detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature, such detail is not needed here. Even with detailed models, ${\rm d}B / {\rm d}t$ always comes down to simple subtraction of income and losses.

@@

@@

The functional-balance assumptions are well-supported by data, at least to first approximation. Fig. \ref{fig:assumptions} shows data sourced from the Biomass and Allometry Database (\textsc{baad}) \citep{Falster-2015b}, which includes records for 21084 individuals on 656 species for various size metrics of individual plants. Dashed lines indicate the slope of predictions under functional-balance assumptions in Table \ref{tab:allometry}a. As is evident, the lines fitted within each species roughly align with predicted functional-balance assumption.

@@

The most important prediction arising from the eqns in Table \ref{tab:allometry} is that the amount of live tissue supporting each unit of leaf area increases with plant height. We can derive an explicit prediction in relation to the assumed allocation model:
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Note that $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are traits, i.e. properties that are assumed approximately constant through ontogeny, so that their product is a constant. Again, the predicted relationship is supported by available data from the \textsc{baad} (Fig. \ref{fig:support_costs}).

@@

Define as \textsc{wplcp}

@@

\begin{itemize}
  \item \citet{Ishihara-2016}
  \item  \citet{Enquist-2007}: sla and wd increase mass-production, where as in our theory effects on mass production are negative. (also contradicts earlier model from \citet{Enquist-1999}, where argued wood density had no effect on mass production)
  \item \citet{Ryan-2006}: reviewed evidence for hydraulic limitation of tree height. Designed to explain decrease in height growth rate for trees. Found mixed support: some species experience decreases in physiological function with height, especially very large species \citep{Koch-2004}, but certainly not not all species \citep{Ryan-2006}. Important to emphasise that demographic patterns observed by \citep{Ryan-2006} (decreases in height growth rate, no decrease in biomass growth) predicted from theory here, without invoking hydraulic limitation. So hydraulic limitation may amplify size related effects but not root cause.
  \item \citet{King-2005} Section 3.3 of King  shows that the cost of turning over leaves, branches and fine roots may halve the height growth rate that could otherwise be attained. \citet{King-1999} also predicts a change in the SLA optimising growth rate, via a cross correlation with self shading. Basically, the much greater path length from roots to leaves in trees vs. seedlings results in a much greater increment in support cost per increment in crown area for trees. This explanation is treated by the growth model of \citet{King-1999}, which predicts that the SLA maximizing growth rate declines as seedlings grow larger, even before the onset of leaf turnover.
  \item \citet{King-1999}:
  \item \citet{Li-2014}:
  \item \citep{Givnish-1988,Yokozawa-1995,Makela-1997,King-2011,Moorcroft-2001}
  \item \citep{Coomes-2011}:
\end{itemize}

@@

In what follows, the assumption of exact isometry is non-critical, what matters is that the amount of live biomass (i.e. excluding heartwood) needed to support an extra unit of leaf area increases with height (as Fig. \ref{fig:support_costs}).  Note that similar increases in construction and maintenance costs with plant size are embedded within various models of plant function \citep[e.g.][]{Givnish-1988, Makela-1997}.

@@

\begin{figure}[ht]
\centering
\includegraphics{main/assumptions.pdf}
\caption{\textbf{Key assumptions of a functional balance model for plant construction}. Data are from the \textsc{baad} \citep{Falster-2015b}. Each dot is a single plant. Lines show standardised major axis lines fitted by species, with intensity of shading adjusted according to strength of the relationship. Dashed lines indicate slope of predictions under functional-balance assumptions in Table \ref{tab:allometry}. 
\label{fig:assumptions}}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics{main/support_costs.pdf}
\caption{\textbf{Amount of live stem tissue per unit leaf area increases with plant height}. Data are from the \textsc{baad} \citep{Falster-2015b}. Each dot is a single plant. Plants \textless{}  1m tall were assumed to have no heartwood. Lines show standardised major axis lines fitted by species, with intensity of shading adjusted according to strength of the relationship. Dashed lines indicate slope of 1, being the prediction of the model in Table \ref{tab:allometry}. The dark line shows the relationship assumed in this paper.
\label{fig:support_costs}}
\end{figure}
\newpage

\begin{figure}[ht]
\centering
\includegraphics{main/growth_light_height.pdf}
\caption{\textbf{Effects of some traits on height growth changes with plant size.} Panels show predicted relationship between specific trait and height growth rate, for a plant of specified height and under a range of shading environments.
\label{fig:growth_light_height}}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics{main/max_leaf_above.pdf}
\caption{\textbf{The effect of traits on growth changes with size and
light environment.} Traits moderate the responsiveness of growth to changes in light
environmentThis figure needs to show more of key result with respect to to size.
Possibly separate figure. \label{fig:shifts}}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics{main/max_leaf_above.pdf}
\caption{\textbf{Low construction cost leads to shade intolerance,
because of costs of high turnover.} Panels show effect of traits on
maximum amount of shading that can be endured before net production (eqn
\ref{eq:dbdt}) reaches zero. Lines indicate relationship for plants of a
given height. \label{fig:wplcp}}
\end{figure}

>>>>

We assume growth is fundamentally driven by energy income and subsequent distribution throughout the plant. According to a standard model, the amount of biomass available for growth, ${\rm d}B / {\rm d}t$ is given by the difference between income (photosynthesis) and losses (respiration and turnover) within the plant \citep{Makela-1997, Thornley-2000} (Fig. \ref{fig:conceptual}a, eqn \ref{eq:dbdt}). Photosynthesis is proportional to the average photosynthetic rate per unit leaf area $\bar{p}$ and total leaf area, $A_{\rm l} = M_{\rm l} / \phi$, where $\phi$ is leaf mass per leaf area. The constant $y$ in eqn \ref{eq:dbdt} converts mol CO$_2$ into units of biomass. We assume that $\bar{p}$ increases with light but with diminishing returns, and respiration and turnover rates of different tissues are constants that might differ by trait (see below). While many vegetation models use a detailed physiological model for calculating $\bar{p}$ and $r_{\rm i}$, e.g. as functions of temperature, such detail is not needed here. Even with detailed models, ${\rm d}B / {\rm d}t$ always comes down to simple subtraction of income and losses.

@@

\subsection{Shade tolerance}
A useful measure of a plant's shade tolerance is the point at which photosynthetic gains just balance the costs of tissue turnover and respiration \citep{Givnish-1988,Baltzer-2007,Lusk-2013}, sometimes referred to as the the ``whole-plant light compensation point'' (\wplcp{}).  In general, the average assimilation rate per leaf area $\bar{p}$ is expected to increase with light level or canopy openness, $E$. The \wplcp can be estimated by solving the value of $E=E^*$ where $\frac{{\rm d}B} {{\rm d}t}(E)$ from eqn \ref{eq:dbdt} is zero, which occurs when:
\begin{equation}\label{eq:wplcp}
\bar{p}(E^*) =\frac{\sum \, M_{\rm i} \left(k_{\rm i} y + r_{\rm i}\right)}{A_{\rm l}}.
\end{equation}
Here $\bar{p}(E)$ is the rate of photosynthesis at canopy openness $E$. Below we express

@@

The functional-balance assumptions are well-supported by data, at least to first approximation. Fig. \ref{fig:assumptions} shows data sourced from the Biomass and Allometry Database (\textsc{baad}) \citep{Falster-2015b}, which includes records for 21084 individuals on 656 species for various size metrics of individual plants. Dashed lines indicate the slope of predictions under functional-balance assumptions in Table \ref{tab:allometry}a. As is evident, the lines fitted within each species roughly align with predicted functional-balance assumption Fig. \ref{fig:assumptions}a-c.

@@

The most important prediction arising from the eqns in Table \ref{tab:allometry} is that the amount of live tissue supporting each unit of leaf area increases with plant height. We can derive an explicit prediction in relation to the assumed allocation model:
\begin{equation}\label{eq:msml}
\frac{M_{\rm b} + M_{\rm s}}{A_{\rm l}} = (1+ \alpha_{\rm b1}) \,\theta \, \rho \, \eta_c \, H.
\end{equation}
Note that $\alpha_{\rm b1}, \theta, \rho, \eta_c$ are traits, i.e. properties that are assumed approximately constant through ontogeny, so that their product is a constant. Again, the predicted relationship is supported by available data from the \textsc{baad} (Fig. \ref{fig:assumptions}d).

@@

@@

\begin{itemize}
  \item \citet{Ishihara-2016}
  \item  \citet{Enquist-2007}: sla and wd increase mass-production, where as in our theory effects on mass production are negative. (also contradicts earlier model from \citet{Enquist-1999}, where argued wood density had no effect on mass production)
  \item \citet{Ryan-2006}: reviewed evidence for hydraulic limitation of tree height. Designed to explain decrease in height growth rate for trees. Found mixed support: some species experience decreases in physiological function with height, especially very large species \citep{Koch-2004}, but certainly not all species \citep{Ryan-2006}. Important to emphasise that demographic patterns observed by \citep{Ryan-2006} (decreases in height growth rate, no decrease in biomass growth) predicted from theory here, without invoking hydraulic limitation. So hydraulic limitation may amplify size related effects but not root cause.
  \item \citet{King-2005} Section 3.3 of King  shows that the cost of turning over leaves, branches and fine roots may halve the height growth rate that could otherwise be attained. \citet{King-1999} also predicts a change in the SLA optimising growth rate, via a cross correlation with self shading. Basically, the much greater path length from roots to leaves in trees vs. seedlings results in a much greater increment in support cost per increment in crown area for trees. This explanation is treated by the growth model of \citet{King-1999}, which predicts that the SLA maximizing growth rate declines as seedlings grow larger, even before the onset of leaf turnover.
  \item \citet{King-1999}:
  \item \citet{Li-2014}:
  \item \citep{Givnish-1988,Yokozawa-1995,Makela-1997,King-2011,Moorcroft-2001}
  \item \citep{Coomes-2011}:
\end{itemize}

@@

In what follows, the assumption of exact isometry is non-critical, what matters is that the amount of live biomass (i.e. excluding heartwood) needed to support an extra unit of leaf area increases with height (as Fig. \ref{fig:assumptions}d).  Note that similar increases in construction and maintenance costs with plant size are embedded within various models of plant function \citep[e.g.][]{Givnish-1988, Makela-1997}.

@@

\begin{figure}[ht]
\centering
\includegraphics{main/assumptions.pdf}
\caption{\textbf{Key assumptions of a functional balance model for plant construction}. Data are from the \textsc{baad} \citep{Falster-2015b}. Each dot is a single plant. Lines show standardised major axis lines fitted by species, with intensity of shading adjusted according to strength of the relationship. Dashed lines indicate slope of predictions under functional-balance assumptions in Table \ref{tab:allometry}. 
The dark line shows the relationship assumed in this paper.
\label{fig:assumptions}}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics{main/growth_light_height.pdf}
\caption{\textbf{Three different responses of growth rate to changes in size, with traits.} Panels show predicted relationship between specific trait and height growth rate, for a plant of specified height and under a range of shading environments.
\label{fig:growth_light_height}}
\end{figure}

% \newpage

% \begin{figure}[ht]
% \centering
% \includegraphics{main/max_leaf_above.pdf}
% \caption{\textbf{The effect of traits on growth changes with size and
% light environment.} Traits moderate the responsiveness of growth to changes in light
% environmentThis figure needs to show more of key result with respect to size.
% Possibly separate figure. \label{fig:shifts}}
% \end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics{main/max_leaf_above.pdf}
\caption{\textbf{Effect of traits on shade tolerance.} Panels show effect of traits on maximum amount of shading that can be endured before net production (eqn \ref{eq:dbdt}) reaches zero. Lines indicate relationship for plants of a given height. \label{fig:wplcp}}
\end{figure}
